Component response rate variation underlies the stability of complex systems

### 2020-03-11 notebook/ms.Rmd
Smooth the text to compare matrices rather than variances in component response rate (which was a bit more difficult to read

**Numerical simulations of random $\mathbf{M}$ with and without $\mathbf{\sigma^{2}_{\gamma}}$**. I used numerical simulations and eigenanalysis to test how variation in $\gamma$ affects stability in random matrices with known properties, comparing the stability of $\textbf{A}$ versus $\mathbf{M} = \gamma\mathbf{A}$. Values of $\gamma$ were sampled from a uniform distribution where $\gamma_{i} \sim \mathcal{U}(0, 2)$ and $\sigma^{2}_{\gamma} = 1/3$ (see Supplementary Information for other distributions, which give similar results). Diagonal elements of simulated $\mathbf{A}$ were standardised to the mean value of sampled $\gamma$ so that $E[d]$ of $\textbf{M}$ and $\textbf{M}$ were guaranteed to be identical. Here I focus on the effect of $\gamma$ across values of $\rho$, and for increasing system sizes ($S$) in random and structured networks. By increasing $S$, the objective is to determine the effect of $\gamma$ as system complexity increases toward the boundary at which stability is realistic for a finite system.

@@

Ten thousand numerical simulations of $\mathbf{M}$ revealed that $\sigma^{2}_{\gamma} > 0$ results in a nonlinear relationship between $\rho$ and $\Re(\lambda_{max})$, which can sometimes increase the stability of the system. Figure 2 shows a comparison of $\Re(\lambda_{max})$ across $\rho$ values when $\sigma^{2}_{\gamma} = 0$ versus $\sigma^{2}_{\gamma} = 1/3$ given $S = 25$, $C = 1$, and $\sigma = 0.4$. For $-0.4 \leq \rho \leq 0.7$ (shaded region of Fig. 2), expected $\Re(\lambda_{max})$ was lower given $\sigma^{2}_{\gamma} = 1/3$. For $\rho \geq -0.1$, the range of $\Re(\lambda_{max})$ values observed given $\sigma^{2}_{\gamma} = 1/3$ also had a lower bound, which included negative values for $\rho = -0.1$ and $\rho = 0$. Hence, across a wide range of system correlations, variation in the response rate of system components had a stabilising effect.

The stabilising effect of $\sigma^{2}_{\gamma}$ across $\rho$ increased with increasing $S$. Figure 3 shows 10000 numerical simulations of $\mathbf{M}$ for $\sigma^{2}_{\gamma} = 0$ versus $\sigma^{2}_{\gamma} = 1/3$ across increasing $S$ given $C = 1$ and $\sigma = 0.2$ ($\sigma$ has been lowered here to better illustrate the effect of $S$; note that now given $S = 25$, $1 = \sigma\sqrt{SC}$). For relatively small systems ($S \leq 25$), positive $\sigma^{2}_{\gamma}$ never decreased the expected $\Re(\lambda_{max})$. But as $S$ increased, the curvilinear relationship between $\rho$ and $\Re(\lambda_{max})$ resulted in $\mathbf{M}$ with lower expected $\Re(\lambda_{max})$ for low magnitudes of $\rho$. In turn, as $S$ increased, and systems became more complex, positive $\sigma^{2}_{\gamma}$ increased the proportion of numerical simulations that were observed to be stable (see below).

@@

**Simulation of random $\mathbf{M}$ across $\mathbf{S}$**. To investigate the effect of $\sigma^{2}_{\gamma}$ on stability across systems of varying complexity, I simulated random $\mathbf{M}$ matrices at $\sigma = 0.4$ and $C = 1$ across $S = \{2, 3, ..., 49, 50\}$. One million $\mathbf{M}$ were simulated for each $S$, and the stability of $\mathbf{M}$ was assessed given $\gamma = 1$ ($\sigma^{2}_{\gamma} = 0$) versus $\gamma \sim \mathcal{U}(0, 2)$ ($\sigma^{2}_{\gamma} = 1/3$). For all $S > 10$, I found that the number of stable random systems was higher given $\sigma^{2}_{\gamma} > 0$ than when $\sigma^{2}_{\gamma} = 0$ (Fig. 3; see Supplementary Information for full table of results), and that the difference between the probabilities of observing a stable system increased with an increase in $S$. In other words, the potential for $\sigma^{2}_{\gamma}$ to affect stability increased with system complexity and was most relevant for systems on the cusp of being too complex to be realistically stable. For the highest values of $S$, nearly all systems that were stable given varying $\gamma$ would not have been stable given $\gamma = 1$.

I also simulated 100000 $\mathbf{M}$ for three types of random networks that are typically interpreted as differing interspecific ecological interaction types [@Allesina2011; @Allesina2012]. These interactions types are competitive, mutualist, or predator-prey, as modelled by off-diagonal elements of $\mathbf{M}$ that are constrained to be negative, positive, or paired such that if $A_{ij} > 0$ then $A_{ji} < 0$, respectively [@Allesina2012] (but are otherwise identical to the purely random $\mathbf{M}$). As $S$ increased, a higher number of stable systems were observed given $\sigma^{2}_{\gamma}>0$ versus $\sigma^{2}_{\gamma}=0$ for competitor and predator-prey, but not mutualist, systems. A higher number of stable systems were observed whenever $S > 12$ and $S > 40$ for competitive and predator-prey systems, respectively (note that $\rho < 0$ by definition for predator-prey systems, making stability more likely overall). The stability of mutualist systems was never affected by $\sigma^{2}_{\gamma}$.

@@

**Simulation of structured $\mathbf{M}$ across $\mathbf{S}$**. To investigate how $\sigma^{2}_{\gamma}$ affects the stability of commonly observed network structures, I simulated 1 million $\textbf{M}$ for small-world [@Watts1998], scale-free [@Albert2002], and cascade food web [@Solow1998; @Williams2000] networks. In all of these networks, rules determining the presence or absence of an interaction between components $i$ and $j$ constrained the overall structure of the network. In small-world networks, interactions between components are constrained so that the expected degree of separation between any two components increases in proportion to $\log(S)$ [@Watts1998]. In scale-free networks, the number of components with which a focal component interacts follows a power law; typically a few components have many interactions while most components have few interactions [@Albert2002]. In cascade food webs, species are ranked and interactions are constrained such that a species $i$ can only feed on $j$ if the rank of $i > j$.

Network structure did not strongly modulate the effect that $\sigma^{2}_{\gamma}$ had on stability. For comparable magnitudes of complexity, structured networks still had a higher number of stable $\mathbf{M}$ for $\sigma^{2}_{\gamma}>0$ compared to $\sigma^{2}_{\gamma}=0$. For random networks, positive $\sigma^{2}_{\gamma}$ increased stability given $S > 10$ ($\sigma = 0.4$ and $C = 1$), and therefore complexity $\sigma \sqrt{SC} \gtrapprox 1.26$. This threshold of complexity was comparable for small-world networks, and slightly lower for scale-free networks (note that algorithms for generating small-world and scale-free networks necessarily led to varying $C$, see methods). Varying $\gamma$ increased stability in cascade food webs for $S > 27$, and therefore at a relatively low complexity magnitudes compared to random predator-prey networks. Overall, network structure does not greatly change the effect that $\sigma^{2}_{\gamma}$ has on increasing the upper bound of complexity within which stability might reasonably be observed.

@@

**System feasibility given $\mathbf{Var(\gamma)}$** For complex systems in which individual system components represent the density of some tangible quantity, it is relevant to consider the feasibility of the system. Feasibilility assumes that values of all components are positive at equilibrium [@Grilli2017; @Dougoud2018; @Song2018]. This is of particular interest for ecological communities because population density ($N$) cannot take negative values, meaning that ecological systems need to be feasible for stability to be biologically realistic [@Dougoud2018]. While my results are intended to be general to all complex systems, and not restricted to species networks, I have also performed a feasibility analysis on all matrices $\mathbf{M}$ tested for stability, and additionally for specific types of ecological communities [@Allesina2012] (e.g., competitive, mutualist, predator-prey; see Supplementary Information). I emphasise that $\gamma$ is not interpreted as population density in this analysis, but instead as a fundamental property of species life history such as expected generation time. Feasibility was unaffected by $\sigma^{2}_{\gamma}$ and instead occurred with a fixed probability of $1/2^{S}$, consistent with a recent proof by Serv&aacute;n et al. [@Servan2018] (see Supplementary Information). Hence, for pure interacting species networks, variation in component response rate (i.e., species generation time) does not affect stability at biologically realistic species densities.

>>>>

**Numerical simulations of random systems with and without $\mathbf{\sigma^{2}_{\gamma}}$**. I used numerical simulations and eigenanalysis to test how variation in $\gamma$ affects stability in random matrices with known properties, comparing the stability of $\textbf{A}$ versus $\mathbf{M} = \gamma\mathbf{A}$. Values of $\gamma$ were sampled from a uniform distribution where $\gamma_{i} \sim \mathcal{U}(0, 2)$ and $\sigma^{2}_{\gamma} = 1/3$ (see Supplementary Information for other $\gamma$ distributions, which gave similar results). Diagonal elements of simulated $\mathbf{A}$ were standardised to the mean value of sampled $\gamma$ so that $E[d]$ of $\textbf{M}$ and $\textbf{M}$ were guaranteed to be identical. Here I focus on the effect of $\gamma$ across values of $\rho$, and for increasing system sizes ($S$) in random and structured networks. By increasing $S$, the objective is to determine the effect of $\gamma$ as system complexity increases toward the boundary at which stability is realistic for a finite system.

@@

Ten thousand numerical simulations of $\mathbf{M}$ revealed that $\sigma^{2}_{\gamma}$ results in a nonlinear relationship between $\rho$ and $\Re(\lambda_{max})$, which can sometimes increase the stability of the system. Figure 2 shows a comparison of $\Re(\lambda_{max})$ across $\rho$ values for $\mathbf{A}$ ($\sigma^{2}_{\gamma} = 0$) versus $\mathbf{M}$ ($\sigma^{2}_{\gamma} = 1/3$) given $S = 25$, $C = 1$, and $\sigma = 0.4$. For $-0.4 \leq \rho \leq 0.7$ (shaded region of Fig. 2), expected $\Re(\lambda_{max})$ was lower in $\mathbf{M}$ than $\mathbf{A}$. For $\rho \geq -0.1$, the lower bound of the range of $\Re(\lambda_{max})$ values also decreased given $\sigma^{2}_{\gamma}$, resulting in negative $\Re(\lambda_{max})$ in $\mathbf{M}$ for $\rho = -0.1$ and $\rho = 0$. Hence, across a wide range of system correlations, variation in the response rate of system components had a stabilising effect.

The stabilising effect of $\sigma^{2}_{\gamma}$ across $\rho$ increased with increasing $S$. Figure 3 shows 10000 numerical simulations of $\mathbf{M}$ across increasing $S$ given $C = 1$ and $\sigma = 0.2$ ($\sigma$ has been lowered here to better illustrate the effect of $S$; note that now given $S = 25$, $1 = \sigma\sqrt{SC}$). For relatively small systems ($S \leq 25$), $\sigma^{2}_{\gamma}$ never decreased the expected $\Re(\lambda_{max})$. But as $S$ increased, the curvilinear relationship between $\rho$ and $\Re(\lambda_{max})$ decreased expected $\Re(\lambda_{max})$ for $\mathbf{M}$ given low magnitudes of $\rho$. In turn, as $S$ increased, and systems became more complex, $\sigma^{2}_{\gamma}$ increased the proportion of numerical simulations that were observed to be stable (see below).

@@

**Simulation of random $\mathbf{M}$ across $\mathbf{S}$**. To investigate the effect of $\sigma^{2}_{\gamma}$ on stability across systems of varying complexity, I simulated random $\mathbf{M = \gamma A}$ matrices at $\sigma = 0.4$ and $C = 1$ across $S = \{2, 3, ..., 49, 50\}$. One million $\mathbf{M}$ were simulated for each $S$, and the stability of $\mathbf{A}$ vesus $\mathbf{M}$ was assessed given $\gamma_{i} \sim \mathcal{U}(0, 2)$ ($\sigma^{2}_{\gamma} = 1/3$). For all $S > 10$, I found that the number of stable random systems was higher in $\mathbf{M}$ than $\mathbf{A}$ (Fig. 3; see Supplementary Information for full table of results), and that the difference between the probabilities of observing a stable system increased with an increase in $S$. In other words, the potential for $\sigma^{2}_{\gamma}$ to affect stability increased with system complexity and was most relevant for systems on the cusp of being too complex to be realistically stable. For the highest values of $S$, nearly all systems that were stable given varying $\gamma$ would not have been stable given $\gamma = 1$.

I also simulated 100000 $\mathbf{M = \gamma A}$ for three types of random networks that are typically interpreted as differing interspecific ecological interaction types [@Allesina2011; @Allesina2012]. These interactions types are competitive, mutualist, or predator-prey, as modelled by off-diagonal elements that are constrained to be negative, positive, or paired such that if $A_{ij} > 0$ then $A_{ji} < 0$, respectively [@Allesina2012] (but are otherwise identical to the purely random $\mathbf{A}$). As $S$ increased, a higher number of stable $\mathbf{M}$ relative to $\mathbf{A}$ were observed for competitor and predator-prey, but not mutualist, systems. A higher number of stable systems were observed whenever $S > 12$ and $S > 40$ for competitive and predator-prey systems, respectively (note that $\rho < 0$ for predator-prey systems, making stability more likely overall). The stability of mutualist systems was never affected by $\sigma^{2}_{\gamma}$.

@@

**Simulation of structured $\mathbf{M}$ across $\mathbf{S}$**. To investigate how $\sigma^{2}_{\gamma}$ affects the stability of commonly observed network structures, I simulated 1 million $\mathbf{M = \gamma A}$ for small-world [@Watts1998], scale-free [@Albert2002], and cascade food web [@Solow1998; @Williams2000] networks. In all of these networks, rules determining the presence or absence of an interaction between components $i$ and $j$ constrained the overall structure of the network. In small-world networks, interactions between components are constrained so that the expected degree of separation between any two components increases in proportion to $\log(S)$ [@Watts1998]. In scale-free networks, the number of components with which a focal component interacts follows a power law; typically a few components have many interactions while most components have few interactions [@Albert2002]. In cascade food webs, species are ranked and interactions are constrained such that a species $i$ can only feed on $j$ if the rank of $i > j$.

Network structure did not strongly modulate the effect that $\sigma^{2}_{\gamma}$ had on stability. For comparable magnitudes of complexity, structured networks still had a higher number of stable $\mathbf{M}$ than $\mathbf{A}$. For random networks, $\sigma^{2}_{\gamma}$ increased stability given $S > 10$ ($\sigma = 0.4$ and $C = 1$), and therefore complexity $\sigma \sqrt{SC} \gtrapprox 1.26$. This threshold of complexity, above which more $\mathbf{M}$ than $\mathbf{A}$ were stable, was comparable for small-world networks, and slightly lower for scale-free networks (note that algorithms for generating small-world and scale-free networks necessarily led to varying $C$; see methods). Varying $\gamma$ increased stability in cascade food webs for $S > 27$, and therefore at a relatively low complexity magnitudes compared to random predator-prey networks ($S > 40$). Overall, network structure did not greatly change the effect that $\sigma^{2}_{\gamma}$ had on increasing the upper bound of complexity within which stability might reasonably be observed.

@@

**System feasibility given $\mathbf{Var(\gamma)}$** For complex systems in which individual system components represent the density of some tangible quantity, it is relevant to consider the feasibility of the system. Feasibilility assumes that values of all components are positive at equilibrium [@Grilli2017; @Dougoud2018; @Song2018]. This is of particular interest for ecological communities because population density ($N$) cannot take negative values, meaning that ecological systems need to be feasible for stability to be biologically realistic [@Dougoud2018]. While my results are intended to be general to all complex systems, and not restricted to species networks, I have also performed a feasibility analysis on all matrices tested for stability. I emphasise that $\gamma$ is not interpreted as population density in this analysis, but instead as a fundamental property of species life history such as expected generation time. Feasibility was unaffected by $\sigma^{2}_{\gamma}$ and instead occurred with a fixed probability of $1/2^{S}$, consistent with a recent proof by Serv&aacute;n et al. [@Servan2018] (see Supplementary Information). Hence, for pure interacting species networks, variation in component response rate (i.e., species generation time) does not affect stability at biologically realistic species densities.
